### 2024-02-20 paper_sisc_first_revision/localpsf.tex
reduced to 32 pages

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. The method computes impulse responses of the operator at scattered points, and interpolates these impulse responses to approximate entries of the integral kernel. To compute impulse responses efficiently, we apply the operator to Dirac combs associated with batches of point sources, which are chosen by solving an ellipsoid packing problem. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are then performed with fast H-matrix methods. This end-to-end procedure is illustrated on a blur problem. We demonstrate the PSF-based method's effectiveness by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and for the initial condition in an advective-diffusive transport problem. While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. The Hessian impulse responses on the other hand typically become more local as the numerical rank increases, which benefits the PSF-based method. Numerical results reveal that the PSF-based preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly $5\times$--$10\times$ reductions in the required number of PDE solves, as compared to classical regularization-based preconditioning and no preconditioning. We also present a comprehensive numerical study for the influence of various parameters (that control the shape of the impulse responses and the rank of the Hessian) on the effectiveness of the advection-diffusion Hessian approximation. The results show that the PSF-based preconditioners are able to form good approximations of high-rank Hessians using only a small number of operator applications.
\end{abstract}

@@

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. By ``non-negative integral kernel,'' we mean that entries of $\Aop$'s integral kernel 
%(defined in Equation \ref{eq:kernel_representation})
are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$. Such operators appear as, for example, Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{BorziSchulz11,Delosreyes15,HinzePinnauUlbrichUlbrich08}, 
%AntilEtAl18
Schur complements in Schur complement methods for solving partial differential equations and Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps)~\cite{ChanMathew94,SmithBjorstadGropp96,ToselliWidlund04}, covariance operators in spatial statistics~\cite{ChenStein21,GentonKeyesTurkiyyah18,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}.

Here, ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a subproblem that involves PDEs, or performing other costly computations. Matrix-free does not mean that matrices are never used; our approximation is ultimately converted into a hierarchical matrix (H-matrix), a compressed matrix format that permits fast and scalable linear algebra operations.

@@

\begin{figure}
	\begin{center}
		%	\begin{subfigure}[b]{0.58\textwidth}
		%		\centering
		\includegraphics[scale=0.35]{kernel_impulse_illustration2.pdf}
		%%		\caption{Kernel and impulse responses}
		%		\label{fig:kernel_impulse_illustration}
		%	\end{subfigure}
		%	\begin{subfigure}[b]{0.41\textwidth}
		%		\centering
		%		\includegraphics[scale=0.55]{frog_ellipsoid_width2.pdf}
		%%		\caption{Impulse response support ellipsoid}
		%		\label{fig:frog_ellipsoid_width}
		%	\end{subfigure}
	\end{center}

	\caption{Left: Matrix created by evaluating the integral kernel $\Phi$ for $\Aop$ (Equation~\ref{eq:kernel_representation}) at all pairs of mesh vertices. This illustration is for the integral kernel given in Equation~\ref{eq:frog_kernel} in Section~\ref{sec:frog}. Dark colors indicate large entries and light colors indicate small entries. Rows and columns are ordered according to a kd-tree hierarchical clustering. Right: Impulse responses
%	$\phi_{x_1}=\Phi(~\cdot~, x_1)$ and $\phi_{x_1}=\Phi(~\cdot~, x_2)$
	associated with points $x_1,x_2\in \Omega$, shown by the two dotted vertical lines. Intuitively, one may think of impulse responses as ``columns'' of the integral kernel.
	}
	\label{fig:frog_kernel_impulse_responses}
\end{figure}

The idea of the proposed PSF-based method is to use \emph{impulse response interpolation} to form a high rank approximation of $\Aop$ using a small number of operator applications. The impulse response, $\phi_x$, associated with a point, $x$, is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure~\ref{fig:batches_intro}). Batches of impulse responses may be thought of intuitively as sets of ``columns'' of the kernel (Figure \ref{fig:frog_kernel_impulse_responses}). To choose the batches, we form ellipsoid estimates for the supports of all $\phi_x$ via a moment method (Figure~\ref{fig:frog_moments_ellipsoid}) that involves applying $\Aop^T$ to a small number of polynomial functions (see Section~\ref{sec:intromoments}). We then use a greedy ellipsoid packing algorithm (Figure~\ref{fig:frog_batches}) to maximize the number of impulse responses per batch.
% which requires us to estimate the supports of the impulse responses $\phi_x$ \emph{before} they are computed. 
Then we interpolate translated and scaled versions of these impulse responses to approximate entries of the operator's integral kernel (Figure~\ref{fig:hmatrix_neighbors_frog}). Adding more batches yields impulse responses at more points, increasing the accuracy of the approximation at the cost of one operator application per batch (Figure~\ref{fig:frog_column_errors}).

\begin{figure}
	\begin{subfigure}[b]{0.65\textwidth}
		\centering
		\includegraphics[scale=0.22]{frog_moments3.pdf}
	\end{subfigure}
	\begin{subfigure}[b]{0.34\textwidth}
		\centering
		\includegraphics[scale=0.44]{frog_ellipsoid_width2.pdf}
		%		\label{fig:frog_ellipsoid_width}
	\end{subfigure}
	\caption{Left: Impulse response moments: scaling factor ($V$), mean ($\mu$), and covariance ($\Sigma$). For each point $x \in \Omega$, the quantity $V(x)$ is the integral of $\phi_x$ over $\Omega$, $\mu(x)$ is the location that $\phi_x$ is centered at, and $\Sigma(x)$ is a matrix with eigenvectors and eigenvalues that characterize the width of $\phi_x$ (see Section~\ref{sec:intromoments}). Right: Ellipsoid support for an impulse response.  This ellipsoid is the set of points within $\tau$ standard deviations of the mean of the Gaussian distribution with mean $\spatialmean(x)$ and covariance $\spatialcov(x)$. The scaling factor $V(x)$ characterizes the magnitude of $\phi_x$.
	}
	\label{fig:frog_moments_ellipsoid}
\end{figure}

@@

\begin{figure}
	\begin{subfigure}[b]{0.30\textwidth}
		\centering
		\includegraphics[scale=0.35]{frog_column_errors_psf5.png}
		%		\caption{PSF(5)}
		\label{fig:blur_error_psf5}
	\end{subfigure}
	\begin{subfigure}[b]{0.30\textwidth}
		\centering
		\includegraphics[scale=0.35]{frog_column_errors_psf10.png}
		%		\caption{PSF(10)}
		\label{fig:blur_error_psf10}
	\end{subfigure}
	\begin{subfigure}[b]{0.30\textwidth}
		\centering
		\includegraphics[scale=0.35]{frog_column_errors_psf20.png}
		%		\caption{PSF(20)}
		\label{fig:blur_error_psf20}
	\end{subfigure}
	\begin{subfigure}[b]{0.08\textwidth}
		\centering
		\includegraphics[scale=0.35]{frog_colorbar_edited1.pdf}
		%			\label{fig:blur_colorbar}
	\end{subfigure}

@@

While the PSF-based method in this paper may be used to approximate any operator that has a locally supported non-negative integral kernel, we are primarily motivated by approximation of high-rank Hessians in distributed parameter inverse problems governed by PDEs. In this section, we provide a brief background on this topic, and explain why existing Hessian approximation methods are not satisfactory.

@@

\begin{figure}
	\centering
	{
		%	\setlength{\tabcolsep}{0.1em}
		%	\renewcommand{\arraystretch}{2.0}
		\begin{tabular}{cccc}
			\includegraphics[scale=0.3, valign=c]{frog_ellipsoid_a=1.0.png} & 
			\includegraphics[scale=0.3, valign=c]{frog_1d_a=1.0.png} & 
			\includegraphics[scale=0.3, valign=c]{ricker_ellipsoid_a=0.0.png} & 
			\includegraphics[scale=0.3, valign=c]{ricker_1d_a=0.0.png} \\
			%
			\includegraphics[scale=0.3, valign=c]{frog_ellipsoid_a=20.0.png} & 
			\includegraphics[scale=0.3, valign=c]{frog_1d_a=20.0.png} & 
			\includegraphics[scale=0.3, valign=c]{ricker_ellipsoid_a=0.23.png} & 
			\includegraphics[scale=0.3, valign=c]{ricker_1d_a=0.23.png} \\
			%
			\includegraphics[scale=0.3, valign=c]{frog_ellipsoid_a=27.0.png} & 
			\includegraphics[scale=0.3, valign=c]{frog_1d_a=27.0.png} & 
			\includegraphics[scale=0.3, valign=c]{ricker_ellipsoid_a=0.249.png} & 
			\includegraphics[scale=0.3, valign=c]{ricker_1d_a=0.249.png} 
		\end{tabular}
	}
	\caption{Illustration of the influence of negative numbers in the integral kernel on the robustness of the ellipsoid estimates for the supports of impulse responses. Left two columns: Blur kernel given in Equation \ref{eq:frog_kernel}. Right two columns: Ricker wavelet-type kernel given by $\Aker(y,x) = (1 - a \gamma)\exp\left(-\gamma/2\right)$, where $\gamma=(y-x)^T \Sigma^{-1} (y-x)$, and $\Sigma=\operatorname{diag}\left(0.0025, 0.01\right)$. Ordered from top to bottom, the results are obtained with $a\in\{1.0, 20.0, 27.0\}$ for the left two columns, and $a\in \{0.0, 0.23, 0.249\}$ for the right two columns. Columns 1 and 3: impulse responses with estimated support ellipsoids indicated by the black ellipses. Red and blue represent positive and negative numbers in the integral kernel, respectively. Columns 2 and 4: one-dimensional slice along the horizontal line indicated in the two-dimensional plots. The dashed gray line is at zero.
	}
	\label{fig:robustness_to_negatives}
\end{figure}

If there are negative numbers in the integral kernel (violating Assumption 3), the ellipsoid estimation procedure may incur errors or fail, leading to poor performance or failure of the proposed method. In Figure \ref{fig:robustness_to_negatives} we investigate the robustness of the ellipsoid support estimation procedure to violations of Assumption 3. We study two integral kernel examples, both of which are parameterized by a quantity that controls how negative the kernels are. We make the following observations:
\begin{itemize}
	\item The larger and more numerous the negative numbers are, the more inaccurate the ellipsoid support estimate is.
	\item The further away from the center of the ellipsoid the negative numbers are, the more influence they have on the quality of the ellipsoid support estimate. This is because moment formulas (Equations \ref{eq:define_mean} and \ref{eq:define_cov}) assign more weight to entries in the kernel that are further from the center.
	\item Negative numbers in the integral kernel affect the proposed method more if they are isolated, and less if they are balanced by nearby positive numbers.
	\item As kernels become more negative, the ellipsoid estimation performs well up to a certain threshold that depends on the spatial distribution of negative and positive entries. After that threshold is crossed, the estimation rapidly transitions to performing poorly and ultimately failing.
\end{itemize} 
%The location of the negative numbers also affects the accuracy of the ellipsoid estimates substantially, but in a more complicated manner. 
For the kernel in the left two columns of Figure \ref{fig:robustness_to_negatives}, negative numbers are interspersed with positive numbers, allowing us to include a large amount of negative numbers in the integral kernel before the proposed method fails. For the kernel in the right two columns, the ellipsoid estimate fails with tiny amounts of negative numbers because the negative numbers are far away from the mean and not balanced by positive numbers at similar distances and angles. In the bottom two rows, we see the aforementioned threshold effect, in which the proposed method rapidly transitions from performing reasonably well to performing poorly with only a small change to the integral kernel.

@@

In summary, computing all the necessary ingredients to evaluate kernel entries of the PSF approximation requires a handful of operator applications (e.g., $6+\nbatch$ operator applications in two dimensions, or $10+\nbatch$ operator applications in three dimensions, with $\nbatch$ typically in the range 1--25), plus comparatively cheap additional overhead costs, most notably performing ellipsoid intersection tests while choosing sample point batches. Once these ingredients are computed, no more operator applications (and thus PDE solves) are required, and approximate kernel entries can be evaluated rapidly. Constructing the H-matrix from kernel entries requires a number of elementary operations that scales polylog linearly in $N$. Using the resulting H-matrix to perform further linear algebra operations also scales polylog linearly in $N$, though the details of these costs depend heavily on the type of H-matrix and operation being performed.

@@

In Table~\ref{tab:frog_psf_vs_hodlr_vs_glr} we compare the
computational cost to approximate the blur kernel from
Equation~\eqref{eq:frog_kernel} using the proposed PSF-based method, the
randomized HODLR (hierarchical off diagonal low rank)
method~\cite{Martinsson16,Hartland_2023}, and GLR (global low rank)
approximation using randomized SVD~\cite{HalkoMartinssonTropp11}. For
these results we vary the quantity $L$ to scale the width of the
impulse responses and hence the rank of the operator. For each case,
we calculate the relative error in the approximation of the kernel
measured in the Frobenius norm, $||\boldsymbol{\Phi} -
\widetilde{\boldsymbol{\Phi}}||_\text{Fro} /
||\boldsymbol{\Phi}||_\text{Fro}$ and show the number of operator
applications required to achieve 20\%, 10\%, and 5\% relative error by
each method. The results reveal superior performance of the PSF
method as compared to the HODLR and GLR methods for all cases. We note
that as we increase the rank and decrease the error tolerance, the
performance of the the HODLR and GLR methods deteriorates.

>>>>

% REQUIRED
\begin{abstract}
	We present an efficient matrix-free point spread function (PSF) method for approximating operators that have locally supported non-negative integral kernels. The proposed method computes impulse responses of the operator at scattered points, and interpolates these impulse responses to approximate entries of the integral kernel. To compute impulse responses efficiently, we apply the operator to Dirac combs associated with batches of point sources, which are chosen by solving an ellipsoid packing problem. The ability to rapidly evaluate kernel entries allows us to construct a hierarchical matrix (H-matrix) approximation of the operator. Further matrix computations are then performed with fast H-matrix methods. This end-to-end procedure is illustrated on a blur problem. We demonstrate the proposed method's effectiveness by using it to build preconditioners for the Hessian operator arising in two inverse problems governed by partial differential equations (PDEs): inversion for the basal friction coefficient in an ice sheet flow problem and for the initial condition in an advective-diffusive transport problem. While for many ill-posed inverse problems the Hessian of the data misfit term exhibits a low rank structure, and hence a low rank approximation is suitable, for many problems of practical interest the numerical rank of the Hessian is still large. The Hessian impulse responses on the other hand typically become more local as the numerical rank increases, which benefits the proposed method. Numerical results reveal that the preconditioner clusters the spectrum of the preconditioned Hessian near one, yielding roughly $5\times$--$10\times$ reductions in the required number of PDE solves, as compared to classical regularization-based preconditioning and no preconditioning. We also present a comprehensive numerical study for the influence of various parameters (that control the shape of the impulse responses and the rank of the Hessian) on the effectiveness of the advection-diffusion Hessian approximation. The results show that the proposed method is able to form good approximations of high-rank Hessians using only a small number of operator applications.
\end{abstract}

@@

We present an efficient \emph{matrix-free} point spread function (PSF) method for approximating operators $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^\gdim$ is a bounded domain, and $L^2(\Omega)'$ is the space of real-valued continuous linear functionals on $L^2(\Omega)$. By ``non-negative integral kernel,'' we mean that entries of $\Aop$'s integral kernel 
%(defined in Equation \ref{eq:kernel_representation})
are non-negative numbers; this is not the same as positive semi-definiteness of $\Aop$. Such operators appear as, for example, Hessians in optimization and inverse problems governed by partial differential equations (PDEs)~\cite{BorziSchulz11,Delosreyes15,HinzePinnauUlbrichUlbrich08}, 
%AntilEtAl18
Schur complements in Schur complement methods for solving partial differential equations and Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps)~\cite{ChanMathew94,SmithBjorstadGropp96,ToselliWidlund04}, covariance operators in spatial statistics~\cite{ChenStein21,GentonKeyesTurkiyyah18,GeogaEtAl20,LindgrenRueLindstrom11}, and blurring operators in imaging~\cite{DenisEtAl11,NagyOleary98}. Here, ``matrix-free'' means that we may apply $\Aop$ and its transpose\footnote{Recall that $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(w) = \left(\Aop^T w\right)(u)$ for all $u,w \in L^2(\Omega)$, where $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(w)$ is the result of applying that linear functional to $w \in L^2(\Omega)$, and similar for operations with $\Aop^T$. }, $\Aop^T$, to functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad w \mapsto \Aop^T w,
\end{equation} 
via a black box computational procedure, 
but cannot easily access entries of $\Aop$'s integral kernel. Evaluating the maps in~\eqref{eq:eval_maps} may require solving a subproblem that involves PDEs, or performing other costly computations. 
%Matrix-free does not mean that matrices are never used; our approximation is ultimately converted into a hierarchical matrix (H-matrix), a compressed matrix format that permits fast and scalable linear algebra operations.

@@

\begin{figure}
	\begin{center}
		%	\begin{subfigure}[b]{0.58\textwidth}
		%		\centering
		\includegraphics[scale=0.3]{kernel_impulse_illustration2.pdf}
		%%		\caption{Kernel and impulse responses}
		%		\label{fig:kernel_impulse_illustration}
		%	\end{subfigure}
		%	\begin{subfigure}[b]{0.41\textwidth}
		%		\centering
		%		\includegraphics[scale=0.55]{frog_ellipsoid_width2.pdf}
		%%		\caption{Impulse response support ellipsoid}
		%		\label{fig:frog_ellipsoid_width}
		%	\end{subfigure}
	\end{center}

	\caption{Left: Matrix created by evaluating the integral kernel $\Phi$ for $\Aop$ (Equation~\ref{eq:kernel_representation}) at all pairs of mesh vertices. This illustration is for the integral kernel in Equation~\ref{eq:frog_kernel}. Dark colors indicate large entries and light colors indicate small entries. Rows and columns are ordered according to a kd-tree hierarchical clustering. Right: Impulse responses
%	$\phi_{x_1}=\Phi(~\cdot~, x_1)$ and $\phi_{x_1}=\Phi(~\cdot~, x_2)$
	associated with points $x_1,x_2\in \Omega$, shown by the two dotted vertical lines. Intuitively, one may think of impulse responses as ``columns'' of the integral kernel.
	}
	\label{fig:frog_kernel_impulse_responses}
\end{figure}

The idea of the proposed method is to use \emph{impulse response interpolation} to form a high rank approximation of $\Aop$ using a small number of operator applications. The impulse response, $\phi_x$, associated with a point, $x$, is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(w) = \left(\rho^*,w\right)_{L^2(\Omega)}$ for all $w \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with batches of points scattered throughout the domain (see Figure~\ref{fig:batches_intro}). Batches of impulse responses may be thought of intuitively as sets of ``columns'' of the kernel (Figure \ref{fig:frog_kernel_impulse_responses}). To choose the batches, we form ellipsoid estimates for the supports of all $\phi_x$ via a moment method (Figure~\ref{fig:frog_moments_ellipsoid}) that involves applying $\Aop^T$ to a small number of polynomials (see Section~\ref{sec:intromoments}). We then use a greedy ellipsoid packing algorithm (Figure~\ref{fig:frog_batches}) to maximize the number of impulse responses per batch.
% which requires us to estimate the supports of the impulse responses $\phi_x$ \emph{before} they are computed. 
Then we interpolate translated and scaled versions of these impulse responses to approximate entries of the operator's integral kernel (Figure~\ref{fig:hmatrix_neighbors_frog}). Adding more batches yields impulse responses at more points, increasing the approximation accuracy at the cost of one operator application per batch (Figure~\ref{fig:frog_column_errors}).

\begin{figure}
	\begin{subfigure}[b]{0.65\textwidth}
		\centering
		\includegraphics[scale=0.22]{frog_moments3.pdf}
	\end{subfigure}
	\begin{subfigure}[b]{0.34\textwidth}
		\centering
		\includegraphics[scale=0.44]{frog_ellipsoid_width2.pdf}
		%		\label{fig:frog_ellipsoid_width}
	\end{subfigure}
	\caption{Left: Impulse response moments: scaling factor ($V$), mean ($\mu$), and covariance ($\Sigma$). For each point $x \in \Omega$, the quantity $V(x)$ is the integral of $\phi_x$ over $\Omega$, $\mu(x)$ is the location that $\phi_x$ is centered at, and $\Sigma(x)$ is a matrix with eigenvectors and eigenvalues that characterize the width of the support of $\phi_x$ about $\mu(x)$ (see Section~\ref{sec:intromoments}). Right: Ellipsoid support for an impulse response.  This ellipsoid is the set of points within $\tau$ standard deviations of the mean of the Gaussian distribution with mean $\spatialmean(x)$ and covariance $\spatialcov(x)$. The scaling factor $V(x)$ characterizes the magnitude of $\phi_x$.
	}
	\label{fig:frog_moments_ellipsoid}
\end{figure}

@@

\begin{figure}
	\begin{subfigure}[b]{0.30\textwidth}
		\centering
		\includegraphics[scale=0.32]{frog_column_errors_psf5.png}
		%		\caption{PSF(5)}
		\label{fig:blur_error_psf5}
	\end{subfigure}
	\begin{subfigure}[b]{0.30\textwidth}
		\centering
		\includegraphics[scale=0.32]{frog_column_errors_psf10.png}
		%		\caption{PSF(10)}
		\label{fig:blur_error_psf10}
	\end{subfigure}
	\begin{subfigure}[b]{0.30\textwidth}
		\centering
		\includegraphics[scale=0.32]{frog_column_errors_psf20.png}
		%		\caption{PSF(20)}
		\label{fig:blur_error_psf20}
	\end{subfigure}
	\begin{subfigure}[b]{0.08\textwidth}
		\centering
		\includegraphics[scale=0.32]{frog_colorbar_edited1.pdf}
		%			\label{fig:blur_colorbar}
	\end{subfigure}

@@

While the method proposed in this paper may be used to approximate any operator that has a locally supported non-negative integral kernel, we are primarily motivated by approximation of high-rank Hessians in distributed parameter inverse problems governed by PDEs. In this section, we provide a brief background on this topic, and explain why existing Hessian approximation methods are not satisfactory.

@@

\begin{figure}
	\centering
	{
		%	\setlength{\tabcolsep}{0.1em}
		%	\renewcommand{\arraystretch}{2.0}
		\begin{tabular}{cccc}
			\includegraphics[scale=0.25, valign=c]{frog_ellipsoid_a=1.0.png} & 
			\includegraphics[scale=0.25, valign=c]{frog_1d_a=1.0.png} & 
			\includegraphics[scale=0.25, valign=c]{ricker_ellipsoid_a=0.0.png} & 
			\includegraphics[scale=0.25, valign=c]{ricker_1d_a=0.0.png} \\
			%
			\includegraphics[scale=0.25, valign=c]{frog_ellipsoid_a=20.0.png} & 
			\includegraphics[scale=0.25, valign=c]{frog_1d_a=20.0.png} & 
			\includegraphics[scale=0.25, valign=c]{ricker_ellipsoid_a=0.23.png} & 
			\includegraphics[scale=0.25, valign=c]{ricker_1d_a=0.23.png} \\
			%
			\includegraphics[scale=0.25, valign=c]{frog_ellipsoid_a=27.0.png} & 
			\includegraphics[scale=0.25, valign=c]{frog_1d_a=27.0.png} & 
			\includegraphics[scale=0.25, valign=c]{ricker_ellipsoid_a=0.249.png} & 
			\includegraphics[scale=0.25, valign=c]{ricker_1d_a=0.249.png} 
		\end{tabular}
	}
	\caption{Illustration of the influence of negative numbers in the integral kernel on the robustness of the ellipsoid estimates for the supports of impulse responses. Left two columns: Blur kernel given in Equation \ref{eq:frog_kernel}. Right two columns: Ricker wavelet-type kernel given by $\Aker(y,x) = (1 - a \gamma)\exp\left(-\gamma/2\right)$, where $\gamma=(y-x)^T \Sigma^{-1} (y-x)$, and $\Sigma=\operatorname{diag}\left(0.0025, 0.01\right)$. Ordered from top to bottom, the results are obtained with $a\in\{1.0, 20.0, 27.0\}$ for the left two columns, and $a\in \{0.0, 0.23, 0.249\}$ for the right two columns. Columns 1 and 3: impulse responses with estimated support ellipsoids indicated by the black ellipses. Red and blue represent positive and negative numbers in the integral kernel, respectively. Columns 2 and 4: one-dimensional slice along the horizontal line indicated in the two-dimensional plots. The dashed gray line is at zero.
	}
	\label{fig:robustness_to_negatives}
\end{figure}

If there are negative numbers in the integral kernel (violating Assumption 3), the ellipsoid estimation procedure may incur errors or fail, leading to poor performance or failure of the proposed method. In Figure \ref{fig:robustness_to_negatives} we investigate the robustness of the ellipsoid support estimation procedure to violations of Assumption 3. We study two integral kernel examples, both of which are parameterized by a quantity that controls how negative the kernels are. We make the following observations:
\begin{itemize}
	\item The larger and more numerous the negative numbers are, the more inaccurate the ellipsoid support estimate is.
	\item The further away from the center of the ellipsoid the negative numbers are, the more influence they have on the quality of the ellipsoid support estimate. This is because moment formulas (Equations \ref{eq:define_mean} and \ref{eq:define_cov}) assign more weight to entries in the kernel that are further from the center.
	\item Negative numbers in the integral kernel affect the proposed method more if they are isolated, and less if they are balanced by nearby positive numbers.
	\item As kernels become more negative, the ellipsoid estimation performs well up to a certain threshold that depends on the spatial distribution of negative and positive entries. After that threshold is crossed, the estimation rapidly transitions to performing poorly and ultimately failing.
\end{itemize} 
%The location of the negative numbers also affects the accuracy of the ellipsoid estimates substantially, but in a more complicated manner. 
For the kernel in the left two columns of Figure \ref{fig:robustness_to_negatives}, negative numbers are interspersed with positive numbers, allowing us to include a large amount of negative numbers before the proposed method fails. For the kernel in the right two columns, the ellipsoid estimate fails with tiny amounts of negative numbers because the negative numbers are far away from the mean and not balanced by positive numbers at similar distances and angles. In the bottom two rows, we see the aforementioned threshold effect, in which the proposed method rapidly transitions from performing reasonably well to performing poorly with only a small change to the integral kernel.

@@

In summary, computing all the necessary ingredients to evaluate kernel entries of the PSF approximation requires a handful of operator applications (e.g., $6+\nbatch$ operator applications in two dimensions, or $10+\nbatch$ operator applications in three dimensions, with $\nbatch$ typically in the range 1--25), plus comparatively cheap additional overhead costs, most notably performing ellipsoid intersection tests while choosing sample point batches. Once these ingredients are computed, no more operator applications (and thus PDE solves) are required, and approximate kernel entries can be evaluated rapidly. Constructing the H-matrix from kernel entries requires a number of elementary operations that scales polylog linearly in $N$. Using the H-matrix to perform linear algebra operations also scales polylog linearly in $N$, though the details of these costs depend heavily on the type of H-matrix and operation being performed.

@@

In Table~\ref{tab:frog_psf_vs_hodlr_vs_glr} we compare the
computational cost to approximate the blur kernel from
Equation~\eqref{eq:frog_kernel} using the proposed PSF-based method, the
randomized HODLR (hierarchical off diagonal low rank)
method~\cite{Martinsson16,Hartland_2023} with 8 hierarchical levels, and GLR (global low rank)
approximation using double-pass randomized SVD~\cite{HalkoMartinssonTropp11}. For
these results we vary the quantity $L$ to scale the width of the
impulse responses and hence the rank of the operator. For each case,
we calculate the relative error in the approximation of the kernel
measured in the Frobenius norm, $||\boldsymbol{\Phi} -
\widetilde{\boldsymbol{\Phi}}||_\text{Fro} /
||\boldsymbol{\Phi}||_\text{Fro}$ and show the number of operator
applications required to achieve 20\%, 10\%, and 5\% relative error by
each method. The results reveal superior performance of the PSF
method as compared to the HODLR and GLR methods for all cases. We note
that as we increase the rank and decrease the error tolerance, the
performance of the the HODLR and GLR methods deteriorates.
